---
title: "Predict movie ratings with MovieLens dataset"
author: "Bart Boerman"
date: "20-12-2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(dplyr) # data wrangling
require(lubridate) # date functions
setwd("~/Data-science/MovieLens")
edx <- readRDS("edx.rds")
options(figcap.prefix = "Figure", figcap.sep = ":", figcap.prefix.highlight = "**")
```

## Overview of capstone project 

[Some text about this capstone project.]

## Dataset description

The training dataset provided by HarvardX contains `r dim(edx)[1]` records and `r dim(edx)[2]` columns with:

* `r n_distinct(edx$userId)` users,
* `r n_distinct(edx$movieId)` movies,
* `r n_distinct(edx$title)` titles, let's investigate how this can differ from number of movies,
* `r n_distinct(edx$genres)` unieke values in the genres column, muliple genres may apply to one genre,
* `r n_distinct(edx$rating)`, ratings, in steps of 0.05 from 0 to 5

The data spreads a timespam from `r min(as_datetime(edx$timestamp))` until `r max(as_datetime(edx$timestamp))`.

```{r example, results="show"}
head(edx, 6) %>% knitr::kable(format = "pandoc", caption = "Example rows") 
```


## Descriptive statistics

```{r}

```

## Data wrangling


## Exploratory data analysis








